---
title: "Project 2 — Dataset 2: Air Quality Sensor Readings"
author: "YOUR NAME HERE"
date: today
format:
  html:
    toc: true
    code-fold: true
    theme: flatly
execute:
  echo: true
  warning: false
---

## Introduction

My classmate David Melchor mentioned the UCI Air Quality dataset on the Week 5 Discussion 5A board — hourly pollutant readings from a set of gas sensors in an Italian city, spread across a wide table with one column per pollutant. He didn't attach the file, so I pulled the actual dataset from the UCI Machine Learning Repository myself, since it's the exact one he referenced and it's public.

## Approach

I'll download the raw file from UCI and put it in this repo without changing its structure, so it's still one row per hour with a separate column for every sensor reading. Then I'll bring it into R, combine the date and time into one proper timestamp, and reshape the pollutant columns into long format so each row is one sensor reading at one point in time. This dataset uses `-200` as a stand-in for a missing reading, so I'll deal with that explicitly rather than let it quietly get treated as a real number. Once it's tidy, I'll look at how carbon monoxide levels change over the course of a day and whether there's any relationship between temperature and benzene levels.

## Data source

UCI Machine Learning Repository, "Air Quality" dataset (De Vito et al.), <https://archive.ics.uci.edu/dataset/360/air+quality>. It's hourly averaged readings from five metal-oxide gas sensors plus a reference analyzer, recorded in an Italian city from March 2004 to April 2005. I downloaded the CSV straight from UCI's site.

## Data structure before tidying

The raw file has 9,357 rows (one per hour) and 15 columns. `Date` and `Time` identify when the reading was taken, and the other 13 columns are the actual measurements — five reference-analyzer pollutant readings (`CO(GT)`, `NMHC(GT)`, `C6H6(GT)`, `NOx(GT)`, `NO2(GT)`), five sensor responses (`PT08.S1` through `PT08.S5`), and three weather readings (`T` for temperature, `RH` for relative humidity, `AH` for absolute humidity). It's wide because every pollutant gets its own column instead of the pollutant being a value in a `pollutant` column.

Two things about how UCI published it that I need to deal with before it's usable:

- It's semicolon-delimited with **commas as decimal points** (European formatting), so `2,6` means 2.6.
- Missing readings are recorded as **`-200`**, not a blank cell.

```{r load}
library(tidyverse)

url <- "https://raw.githubusercontent.com/amine-souissi0/data607-project2/main/dataset2_air_quality/data/air_quality.csv"

raw <- read_delim(url, delim = ";", locale = locale(decimal_mark = ","), show_col_types = FALSE)

raw |> select(Date, Time, `CO(GT)`, `NOx(GT)`, T) |> head()
```

`locale(decimal_mark = ",")` tells `read_delim()` to parse `2,6` as the number 2.6 instead of choking on it or reading it as text — that one setting handles the European number format for every numeric column at once.

## Transformation steps

```{r tidy}
tidy <- raw |>
  mutate(datetime = dmy_hms(paste(Date, str_replace_all(Time, "\\.", ":")))) |>
  select(-Date, -Time) |>
  rename_with(~ str_replace_all(.x, "[().]", "_") |>
                 str_replace_all("_+$", "") |>
                 str_to_lower()) |>
  pivot_longer(
    cols      = -datetime,
    names_to  = "sensor",
    values_to = "reading"
  ) |>
  mutate(reading = na_if(reading, -200))

tidy
```

Step by step:

- `dmy_hms(paste(Date, ...))` glues the date and time columns into one real timestamp. The time in the raw file looks like `18.00.00` instead of `18:00:00`, so I swap the periods for colons first with `str_replace_all()`.
- `rename_with()` cleans up all 13 measurement column names at once — `CO(GT)` becomes `co_gt`, `PT08.S1(CO)` becomes `pt08_s1_co`, and so on. That's the "consistent naming convention" step: no more parentheses or dots in a column name.
- `pivot_longer()` is the actual reshape — instead of 13 separate pollutant columns, I get one `sensor` column naming which reading it is, and one `reading` column with the value.
- `na_if(reading, -200)` is the missing-data decision: UCI's own documentation for this dataset says `-200` marks a reading that wasn't recorded, so I convert every `-200` to a real `NA`. Leaving it as `-200` would silently wreck any average I compute later — a -200 mixed into a mean is a disaster, not just a slightly wrong number.

```{r missingness}
missing_by_sensor <- tidy |>
  summarise(pct_missing = round(100 * mean(is.na(reading)), 1), .by = sensor) |>
  arrange(desc(pct_missing))

missing_by_sensor |>
  rename(Sensor = sensor, `% missing` = pct_missing) |>
  knitr::kable()
```

`nmhc_gt` is missing on about 90% of hours — that's not a fluke. UCI's own notes for this dataset say the non-methane hydrocarbon sensor failed partway through the study and stopped producing usable readings for most of it. Everything else is missing somewhere between 4% and 18% of the time, which is a much more normal amount of sensor downtime. **My decision:** I'm keeping `nmhc_gt` in the tidy dataset for completeness, since it's still honest data about that sensor, but I'm leaving it out of the analysis below because 90% missing isn't enough to say anything meaningful about it.

## Analytical methods

Two questions, both worked out from the tidy long table:

1. **Daily pattern** — average carbon monoxide (`co_gt`) for each hour of the day, across the whole 13 months, to see when pollution peaks.
2. **Temperature and benzene** — whether warmer hours tend to have higher or lower benzene (`c6h6_gt`) readings, using a scatterplot and a correlation.

```{r hourly}
hourly_co <- tidy |>
  filter(sensor == "co_gt", !is.na(reading)) |>
  mutate(hour = hour(datetime)) |>
  summarise(avg_co = mean(reading), .by = hour) |>
  arrange(hour)

ggplot(hourly_co, aes(x = hour, y = avg_co)) +
  geom_line(color = "#1F4E79", linewidth = 1) +
  geom_point(color = "#1F4E79") +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  labs(title = "Average CO by hour of day",
       x = "Hour of day (0–23)", y = "CO (mg/m³)") +
  theme_minimal()
```

There are two clear peaks — one around 8–9 a.m. and a bigger one around 7 p.m., with the highest average CO at 7 p.m. (about 3.7 mg/m³) and the lowest around 5 a.m. (about 0.7 mg/m³). That's basically a rush-hour signature: the city's air gets noticeably dirtier right when people are commuting.

```{r temp-benzene}
wide_pair <- tidy |>
  filter(sensor %in% c("t", "c6h6_gt")) |>
  pivot_wider(names_from = sensor, values_from = reading) |>
  drop_na(t, c6h6_gt)

correlation <- round(cor(wide_pair$t, wide_pair$c6h6_gt), 3)

ggplot(wide_pair, aes(x = t, y = c6h6_gt)) +
  geom_point(alpha = 0.15, color = "#1F4E79") +
  geom_smooth(method = "lm", color = "#C0504D", se = FALSE) +
  labs(title = paste0("Temperature vs. benzene (correlation = ", correlation, ")"),
       x = "Temperature (°C)", y = "Benzene, C6H6 (µg/m³)") +
  theme_minimal()
```

The correlation is about 0.2 — a weak positive relationship, not a strong one. Warmer hours lean slightly toward higher benzene, but temperature clearly isn't the main thing driving it; traffic volume and how well the air mixes overnight probably matter a lot more.

## Conclusions

<!-- TODO (student): write this in your own words. Prompts: -->

- What does the rush-hour CO pattern tell you about where this pollution is probably coming from?
- Why does a correlation of about 0.2 count as "weak," and what would need to be true for it to count as strong?
- Was dropping `nmhc_gt` from the analysis (instead of trying to guess the missing values) the right call? What's the argument either way?
- How would you extend this — using the `pt08_s2_nmhc` sensor as a stand-in for the broken NMHC sensor, looking at a specific season, or something else?

## AI Use

Anthropic. (2026). *Claude Sonnet 5* [Large language model]. https://claude.ai. Accessed September 30, 2026.

I used Claude to help plan the tidying pipeline, draft the R code for each step, and check the hourly averages and correlation independently before trusting them. I reviewed the code and can explain what each step does.
